---
title: "Integration"
author:
  - name: "Emir Turkes and Naoto Watamura, UK DRI at UCL"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
link-citations: true
output:
  html_document:
    code_folding: show
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(
    inputFile,
    encoding = encoding,
    output_file = file.path("..", "results", "integration.html")
  )})
---

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 25px; text-indent: 0;
}
div.main-container {max-width: none; width: 100%;}
</style>

```{r}
# Copyright 2025 Emir Turkes, Naoto Watamura, UK DRI at UCL
#
# Licensed under the Apache License, Version 2.0 (the "License");
# you may not use this file except in compliance with the License.
# You may obtain a copy of the License at
#
#     http://www.apache.org/licenses/LICENSE-2.0
#
# Unless required by applicable law or agreed to in writing, software
# distributed under the License is distributed on an "AS IS" BASIS,
# WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
# See the License for the specific language governing permissions and
# limitations under the License.

# Load packages.
# --------------
library(conflicted)
packages <- c("Seurat", "ggplot2", "dplyr", "ggrepel")
invisible(
  suppressPackageStartupMessages(
    lapply(packages, FUN = library, character.only = TRUE)
  )
)
# --------------

# Define functions.
# -----------------
source("utils.R")
# -----------------

# Add paths.
# ----------
data_dir <- file.path("..", "data")
cache_dir <- file.path("..", "cache")
if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}
# ----------

# Global settings.
# ----------------
knitr::opts_chunk$set(fig.width = 7, fig.height = 5, dpi = 96)
# ----------------
```

# Prep Data

```{r}
snrnaseq <- readRDS(file.path(data_dir, "snrnaseq.rds"))
geomx <- readRDS(file.path(data_dir, "geomx.rds"))

geomx$status <- factor(
  ifelse(
    geomx$genotype %in% c("MAPTKI", "P301S"), yes = "Homeostatic",
    no = ifelse(
      geomx$genotype %in% c("NLGF_MAPTKI", "NLGF_P301S"), yes = "Activated",
      no = "Other"
    )
  )
)

snrnaseq@active.ident <- snrnaseq$genotype
snrnaseq <- subset(
  snrnaseq, idents = c("MAPTKI", "P301S", "NLGF_MAPTKI", "NLGF_P301S")
)
snrnaseq$genotype <- factor(
  snrnaseq$genotype,
  levels = c("MAPTKI", "P301S", "NLGF_MAPTKI", "NLGF_P301S")
)

rds <- file.path(cache_dir, "snrnaseq_geneids.rds")
if (file.exists(rds)) {
  snrnaseq_geneids <- readRDS(rds)
} else {
  snrnaseq_geneids <- suppressWarnings(GetAssayData(snrnaseq, assay = "RNA"))
  gene_anno <- data.frame(
    external_gene_name = unlist(snrnaseq@misc$geneids),
    ensembl_gene_id = rownames(
      suppressWarnings(GetAssayData(snrnaseq, assay = "RNA"))
    )
  )
  gene_anno <- gene_anno[
    gene_anno$ensembl_gene_id %in% rownames(snrnaseq_geneids),
  ]
  rownames(snrnaseq_geneids) <- gene_anno$external_gene_name
  rm(gene_anno)
  snrnaseq_geneids <- CreateSeuratObject(
    counts = snrnaseq_geneids, assay = "RNA", meta.data = snrnaseq@meta.data
  )

  snrnaseq_geneids <- SCTransform(snrnaseq_geneids, verbose = FALSE)
  snrnaseq_geneids <- RunPCA(snrnaseq_geneids, npcs = 30, verbose = FALSE)
  snrnaseq_geneids <- RunUMAP(snrnaseq_geneids, dims = 1:18, verbose = FALSE)
  add_df <- data.frame(Embeddings(snrnaseq_geneids, reduction = "umap"))
  snrnaseq_geneids$umap1 <- add_df[ , 1]
  snrnaseq_geneids$umap2 <- add_df[ , 2]

  saveRDS(snrnaseq_geneids, file = rds)
}

rds <- file.path(cache_dir, "snrnaseq_reprocessed.rds")
if (file.exists(rds)) {
  snrnaseq <- readRDS(rds)
} else {
  snrnaseq <- SCTransform(snrnaseq, verbose = FALSE)
  snrnaseq <- RunPCA(snrnaseq, npcs = 30, verbose = FALSE)
  snrnaseq <- RunUMAP(snrnaseq, dims = 1:18, verbose = FALSE)
  add_df <- data.frame(Embeddings(snrnaseq, reduction = "umap"))
  snrnaseq$umap1 <- add_df[ , 1]
  snrnaseq$umap2 <- add_df[ , 2]

  saveRDS(snrnaseq, file = rds)
}

ElbowPlot(snrnaseq, ndims = 30)
red_dim_plot(
  snrnaseq, x = "umap1", y = "umap2", color = "genotype", type = "cat"
)

pca <- geomx[["pca"]]
pca@feature.loadings[ , 2] <- pca@feature.loadings[ , 1]
pca@cell.embeddings[ , 2] <- pca@cell.embeddings[ , 1]
geomx[["pca"]] <- pca

add_df <- data.frame(Embeddings(geomx, reduction = "pca"))
geomx$pca1 <- add_df[ , 1]
geomx$pca5 <- add_df[ , 5]
rm(add_df)

red_dim_plot(geomx, x = "pca1", y = "pca5", color = "genotype", type = "cat")
```

# Integration

```{r}
anchors <- FindTransferAnchors(
  geomx, snrnaseq_geneids,
  normalization.method = "SCT", recompute.residuals = FALSE,
  reference.reduction = "pca", project.query = TRUE,
  dims = 1:2, verbose = FALSE
)
predictions <- MapQuery(
  anchors, snrnaseq_geneids, geomx,
  refdata = "status", reference.dims = 1:2, query.dims = 1:18,
  verbose = FALSE
)
snrnaseq$status <- factor(predictions$predicted.id)

red_dim_plot(snrnaseq, x = "umap1", y = "umap2", color = "status", type = "cat")

snrnaseq$genotype_status <- factor(
  paste(
    snrnaseq$genotype, snrnaseq$status, sep = "_"
  ),
  levels = c(
    "MAPTKI_Activated", "MAPTKI_Homeostatic",
    "P301S_Activated", "P301S_Homeostatic",
    "NLGF_MAPTKI_Activated", "NLGF_MAPTKI_Homeostatic",
    "NLGF_P301S_Activated", "NLGF_P301S_Homeostatic"
  )
)
table(snrnaseq$genotype_status)

red_dim_plot(snrnaseq, x = "umap1", y = "umap2", color = "genotype_status")

snrnaseq$genotype_status_cleaned <- as.character(snrnaseq$genotype_status)

snrnaseq$genotype_status_cleaned[
  grepl("^MAPTKI", snrnaseq$genotype_status_cleaned)
] <- "MAPTKI"
snrnaseq$genotype_status_cleaned[
  grepl("^P301S", snrnaseq$genotype_status_cleaned)
] <- "P301S"

snrnaseq$genotype_status_cleaned <- factor(
  snrnaseq$genotype_status_cleaned,
  levels = c(
    "MAPTKI", "P301S",
    "NLGF_MAPTKI_Activated", "NLGF_MAPTKI_Homeostatic",
    "NLGF_P301S_Activated", "NLGF_P301S_Homeostatic"
  )
)

red_dim_plot(
  snrnaseq, x = "umap1", y = "umap2",
  color = "genotype_status_cleaned", type = "cat"
)

rds <- file.path(cache_dir, "snrnaseq_integrated.rds")
if (!file.exists(rds)) {
  saveRDS(snrnaseq, file = rds)
}
```

# References

This is the concluding section of the document, where we output the `sessionInfo`, and create a bibliography for works cited.

```{r}
sessionInfo()
```
